---
title: "Fatty Acid Retention Time Prediction"
author: "RLionheart"
date: "2/3/2020"
output: html_document
---

This markdown script controls the procedure for predicting the retention times of targeted fatty acids run on the QExactive Mass Spectrometer.

It contains two major sections:
Section I: Import and cleaning/rearranging of data.
Section II: Create predictive fatty acid retention time graphics.


```{r setup, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(plotly)
library(reshape2)
library(tidyverse)
options(scipen=999)

source("src/Functions.R")

processed.folder <- "data_processed" 
figures.folder <- "figures"
intermediate.folder <- "data_intermediate"
dir.create(file.path(".", processed.folder))  
dir.create(file.path(".", figures.folder))  
dir.create(file.path(".", intermediate.folder))  

if (length(dir(path = "data_processed/")) == 0) {
  cat("\nData_processed subdirectory has been created and is ready for new data.")
} else {
  cat("Warning: data_processed subdirectory is not empty. Empty contents before continuing.\n")
}
```


If you would like to empty the data_processed/, data_intermediate, and figures/ subdirectories, run the following code.
```{r, include = TRUE}
toClear <- c("data_processed/", "data_intermediate", "figures/")
f <- list.files(toClear, include.dirs = F, full.names = T, recursive = T)
file.remove(f)

print("Subdirectories emptied.")
```

----------------------------------------------------------------------------------------------------------------------------

Section I: Import all MSDial files that have been split by Area, Mass/charge (Mz), Retention Time (RT), and Signal to Noise (SN).

In the MSDial datasets:
Set header, filter unknowns.
Change variable classes from character/factor to numeric.
Rearrange the dataframes and combine to a single frame in long format.
Standardize dataset by removing "Ingalls_" prefixes from compound names, and removing the syntactically correct "X" from Replicates.

***
Inputs: 
"data_raw/*file.pattern*.csv

Outputs: 
"data_intermediates/MSDial_combined_*file.pattern*_*DATE*.csv"
***

*User action required*
Enter the file.pattern required for your files. 
- The file.pattern variable should be a character string that applies to all the MSDial files from your run.

```{r Pattern matching, include = TRUE}
file.pattern <- "FA"

print(paste("Your file pattern is:", file.pattern))
```

```{r MSDial Imports, include = TRUE}
source("src/File_imports.R")

print("Required files imported.")
```

*User action required*
Enter the existing filenames of your run, reassigning their values to the strings below. 

The "File_imports.R" script assigns the variables in R to their filename in the directory, so if your positive Area file is "data_raw/PositiveArea.csv", it will be imported to this code as PositiveArea. 
Those imported files need to be renamed to the given variables so the rest of the code will know which files to edit for the pipeline steps.

Enter the matching pattern to help the program identify your files. 
This matching pattern needs to be a string that is unique to the file names that are defined just below. 

```{r Dataset reassignment, include = TRUE}
# Dataset variables: 
Area.FA.data <- Area_KM1906_FA_DepthProfiles # Enter your filenames here
Mz.FA.data <- Mz_KM1906_FA_DepthProfiles
RT.FA.data <- RT_KM1906_FA_DepthProfiles
SN.FA.data <- SN_KM1906_FA_DepthProfiles

matching.pattern <- "FA.data"

print(paste(file.pattern, "variables assigned."))
```

Rearrange and export.
```{r Dataset rearrangement, include = TRUE}
source("src/MSDial_rearrange.R")

print("Data rearrange complete.")
```

-----------------------------------------------------------------------------------------------------------------------------------------

Section II: Use fatty acid standard runs to predict shifting retention times in those compounds without standards.

In the Retention Prediction step:
Import files.
Do required data splitting (ie, split by size fractionation or location, depending on the research question.)
Create random "dummy" data to imitate comparison runs.
Visualize real and dummy data together.
Using the compounds that have standards, use the max and min difference between them to create a prediction interval for retention times.


***
Inputs: 
"data_intermediate/MSDial_combined_*file.pattern*_*DATE*.csv"
"data_extras/*<file_with_expected_RTs>*.csv" 

Outputs: 
"figures/Real_and_Fake_RT_*DATE*.png"
***

*User action required*
Ensure that the file containing the expected retention time (defined as FA.expected) is entered properly below.
```{r Import retention time files, include = TRUE}
FA.expected <- read.csv("data_extras/FA_Expected_RT.csv", stringsAsFactors = FALSE) %>%
  rename(Metabolite.Name = Name) %>%
  rename(RT.Expected = RT) %>%
  select(Metabolite.Name, RT.Expected)

source("src/RT_file_import.R")

print("Imported fatty Acid data combined with expected data.")
```

Separate replicates by size fractionation.
```{r Separate size fractions, include = TRUE}
fatty.acid_0.2 <- FA.expected %>%
  filter(!str_detect(Replicate.Name, "0.3")) 

fatty.acid_0.3 <- FA.expected %>%
  filter(!str_detect(Replicate.Name, "0.2")) 

print("Split by size fractionation.")
```


```{r Create random values for experimentation, include = TRUE}
# Create random values
dummy.data <- fatty.acid_0.2 %>%
  group_by(Metabolite.Name, Replicate.Name) %>%
  mutate(Random.RT = runif(1, RT.Value, RT.Value + 1)) %>%
  group_by(Metabolite.Name) %>%
  mutate(Random.Mean = mean(Random.RT)) %>%
  select(Replicate.Name:RT.Expected, Mean.RT.Value, Random.RT, Random.Mean, Cmpd.with.Std)

print("Column with random data added to real data.")
```

First plot of retention times  
```{r Retention time plot, include = TRUE}
real.RT.plot <- ggplot(fatty.acid_0.2, aes(x = Metabolite.Name, y = Mean.RT.Value, fill = Cmpd.with.Std)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(Mean.RT.Value, digits = 2)), 
            position=position_dodge(width=0.9), 
            vjust=-0.25, size = 2.5) +
  theme(axis.text.x = element_text(angle = 90, size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        strip.text = element_text(size = 10)) +
ggtitle("Fatty Acids: 0.2 Size Fraction")

# And with dummy data
dummy.plot <- ggplot(dummy.data, aes(x = Metabolite.Name, y = Random.Mean, fill = Cmpd.with.Std)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("firebrick3", "chartreuse4", "deepskyblue3")) +
  geom_text(aes(label = round(Random.Mean, digits = 2)), 
            position=position_dodge(width=0.9), 
            vjust=-0.25, size = 2.5) +
  theme(axis.text.x = element_text(angle = 90, size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        strip.text = element_text(size = 10)) +
  ggtitle("Random RT Values")

print("Plots created.")
```

Plot to console real and random data together, and save to the figures/ folder.
```{r Combined plots, include = TRUE, fig.align="center", fig.width=9, fig.height=7}
require(gridExtra)

real.and.dummy.plot <- grid.arrange(real.RT.plot, dummy.plot, nrow = 2)

currentDate <- Sys.Date()
plotFileName <- paste("figures/All_Real_and_Fake_RT_", currentDate, ".png", sep = "")

ggsave(plotFileName, plot = real.and.dummy.plot, dpi = 600, width = 10, height = 10, units = "in")

```


Retention Time Tables 
```{r Retention time tables, include = TRUE}
real.standards <- fatty.acid_0.2 %>%
  filter(!str_detect(Replicate.Name, "IS")) %>%
  filter(str_detect(Replicate.Name, "_Std_")) %>%
  filter(str_detect(Metabolite.Name, "_Std")) %>%
  mutate(RT.Diff = RT.Value - RT.Expected) %>% 
  select(Replicate.Name, Metabolite.Name, RT.Expected, RT.Value, Mean.RT.Value:RT.Diff)

dummy.standards <- dummy.data %>%
  filter(!str_detect(Replicate.Name, "IS")) %>%
  filter(str_detect(Replicate.Name, "_Std_")) %>%
  filter(str_detect(Metabolite.Name, "_Std")) %>%
  group_by(Metabolite.Name) %>%
  mutate(Random.RT.Mean = mean(Random.RT)) 

print("Standard runs isolated.")

```

Real + Dummy standards defined 
```{r Define real and dummy standards, include = TRUE}
std.RT.plot <- ggplot(real.standards, aes(x = Metabolite.Name, y = Mean.RT.Value, label = Mean.RT.Value)) +
  geom_bar(stat = "identity", position = "dodge", fill = "cornflowerblue") +
  geom_text(aes(label = round(Mean.RT.Value, digits = 2)), position=position_dodge(width=0.9), vjust=-0.25) +
  theme(axis.text.x = element_text(angle = 90, size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        strip.text = element_text(size = 10)) +
  ggtitle("Fatty Acids: Standard Compounds")

dummy.std.RT.plot <- ggplot(dummy.standards, aes(x = Metabolite.Name, y = Random.Mean)) +
  geom_bar(stat = "identity", position = "dodge", fill = "deepskyblue3") +
  geom_text(aes(label = round(Random.Mean, digits = 2)), position=position_dodge(width=0.9), vjust=-0.25) +
  theme(axis.text.x = element_text(angle = 90, size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        strip.text = element_text(size = 10)) +
  ggtitle("Random Standards")
```

Visualize standard plots.
```{r Combined plots, include = TRUE, fig.align="center", fig.width=9, fig.height=7}
#require(gridExtra)

stds.together <- dummy.standards %>%
  select(Metabolite.Name, Mean.RT.Value, Random.RT.Mean) %>%
  rename(Real.RT.Value = Mean.RT.Value) %>%
  rename(Random.RT.Value = Random.RT.Mean) %>%
  unique()

stds.together <- melt(stds.together) %>%
  rename(RT.Type = variable) %>%
  rename(Retention.Time = value)
stds.together.plot <- ggplot(stds.together, aes(Metabolite.Name, Retention.Time, fill = RT.Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 90, size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        strip.text = element_text(size = 10)) +
  ggtitle("Real vs Random Standards Retention Times")

stds.together.plot.g <- ggplotly(stds.together.plot)
stds.together.plot.g

plotFileName <- paste("figures/Stds_Real_and_Fake_", currentDate, ".png", sep = "")
ggsave(plotFileName, plot = stds.together.plot, dpi = 600, width = 10, height = 10, units = "in")
```

Real + Dummy standards plotted
```{r High and low prediction estimate, include = TRUE}
stds.differences <- stds.together %>%
  group_by(Metabolite.Name) %>%
  mutate(Difference = abs(Retention.Time[RT.Type == "Real.RT.Value"] - Retention.Time[RT.Type == "Random.RT.Value"])) %>%
  select(Metabolite.Name, Difference) %>%
  unique() %>%
  ungroup() %>%
  mutate(Min.diff = min(Difference)) %>%
  mutate(Max.diff = max(Difference)) %>%
  select(Min.diff, Max.diff) %>%
  unique()

high.estimate <- stds.differences$Max.diff
low.estimate <- stds.differences$Min.diff

print("Standard differences created.")
```


Combine estimates and plot
```{r Plot estimates, include = TRUE, fig.align="center", fig.width=15, fig.height=7}
high.low.predictions <- fatty.acid_0.2 %>%
  select(Replicate.Name, Metabolite.Name, RT.Value) %>%
  mutate(High.RT = RT.Value + high.estimate) %>%
  mutate(Low.RT = RT.Value + low.estimate) %>%
  filter(!str_detect(Metabolite.Name, "Std|IS"))

high.low.plot.table <- melt(high.low.predictions) %>%
  rename(RT.Value.Prediction = variable) %>%
  rename(RT.Value = value)


prediction.plot <- ggplot(high.low.plot.table, aes(fill = RT.Value.Prediction, y = RT.Value, x = Metabolite.Name)) + 
  geom_bar(position="dodge", stat="identity") +
  scale_fill_manual(values=c("firebrick3", "navyblue", "skyblue")) +
  theme(axis.text.x = element_text(angle = 90, size = 10, vjust = 0.5),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        strip.text = element_text(size = 10)) + 
  ggtitle("Predicted Fatty Acid Retention Times")
prediction.plot.g <- ggplotly(prediction.plot)
prediction.plot.g

plotFileName <- paste("figures/Predicted_RTs_", currentDate, ".png", sep = "")
ggsave(plotFileName, plot = prediction.plot, dpi = 600, width = 10, height = 10, units = "in")
```

